---
title: "Munge"
author: "YOUR NAME HERE"
date: "Week 03"
format:
  html:
    embed-resources: true
---


# New York City Airbnb

This lab is based on the New York City Airbnb Open Data from Kaggle. I want you to spend some more time using our data visualization and transformation skills to explore a new data set. Since 2008, guests and hosts have used Airbnb to expand on traveling possibilities and present more unique, personalized way of experiencing the world. This data set describes the listing activity and metrics in NYC, NY for 2019.

## Goals:

- Practice more with ggplot [Data Visualisation](https://r4ds.had.co.nz/data-visualisation.html)
- First steps with data transformations [Transformations](https://r4ds.had.co.nz/transform.html)
- Guided EDA [EDA](https://r4ds.had.co.nz/exploratory-data-analysis.html)

```{r}
#| fig-align: 'center'
#| fig-out: '75%'

url <- "https://imgs.xkcd.com/comics/realistic_criteria.png"
knitr::include_graphics(url)
```


# Setup

While not required, starting a new R session will keep your global environment uncluttered. To start a new R session, go to the "Session" entry and then go to "Restart R" which is about half-way down the menu. There is a short-cut you can use of Ctrl+Shift+F10. (Confession, I don't use this short-cut. I can never remember it.)

```{r}
#| fig-align: 'center'
#| fig-out: '75%'

url <- "./includes/restart-r-session.png"
knitr::include_graphics(url)
```

This is especially important in a lab such as tonight's because we are going to use a different data set for the lab.

```{r}
#| label: setup
#| warning: false
library(knitr)
library(rio)
library(tidyverse)
abnb <- import("./data/new-york-city-airbnb-open-data.zip")
```

- The data for this lab is stored in a CSV file.
- That CSV file is stored in a ZIP file to make it smaller & easier to manage.
- Check the global environment & confirm you have loaded the `abnb` data.


# Lab

If you are coming back in, run `setup` before working on the tasks below.

## Data

1. Read about the [New York City Airbnb](https://www.kaggle.com/dgomonov/new-york-city-airbnb-open-data/data#) data.
    - Your first step with any new data set should be to learn as much as you can about the origins, goals, and biases in the data.
    - And figure out the grain of the data.
2. Spend a few minutes looking at the data. The "Environment" pane is on the lower right-hand side of the RStudio Dekstop interface. Click on "abnb" to open the data in a new editor tab. This makes it easier to explore

```{r}
#| fig-align: 'center'
#| fig-out: '75%'

url <- "./includes/rstudio-interface.png"
knitr::include_graphics(url)
```

## Task 01. How many rows of Airbnb data?

- Question: How many rows of data are in the Airbnb, `abnb`, data?
- Answer: 

```{r}
## YOUR CODE HERE!!!dim(abnb)

# OR JUST USE THE GUI
```

You can solve Task 01 without writing any code.

## Task 02. How many columns are in Airbnb?

- Question: How many columns are in the `abnb` data?
- Answer: 

```{r}
## YOUR CODE HERE!!!dim(abnb)

# OR JUST USE THE GUI
```

You can solve Task 02 without writing any code.

## Task 03. What is the grain of this Airbnb data?

- Question: What is the grain of the `abnb` data?
- Answer: AirBNBs in New York City

 To define the "grain" think about what makes each row unique? If you count all the rows of data, what is it you have counted? You should be able to write this out in just a couple of sentences.

- For more information about the data, please read the [Kaggle documentation](https://www.kaggle.com/dgomonov/new-york-city-airbnb-open-data/data#).
- If you aren't familiar with Airbnb, look at [their website](https://www.airbnb.com).
- If you want to read more about data graininess, [Data Granularity](https://en.wikipedia.org/wiki/Granularity#Data_granularity)

## Task 04. Number of B&Bs are there per `neighbourhood_group`. 

- Question: Which neighborhood has the most B&Bs?
- Answer:

Task: Draw a bar plot showing the number of B&Bs per `neighbourhood_group` to answer the question.

```{r}
## YOUR ANSWER HERE!
abnb |> count(neighbourhood_group)
```

## Task 05. How many B&Bs are in Queens?

- Question: How many B&Bs are in Queens?
- Answer: 5,666

Task: Use the dplyr verbs we learned this evening to answer the question. Replace my question marks with code to find the answer.

- Don't try to use your bar plot and guess. Transform the data!
- This will change the grain of the data.

```{r}
## Three solutions!
abnb |> filter(neighbourhood_group == "Queens")

abnb |>
  group_by(neighbourhood_group) |>
  summarize(n = n())

abnb |>
  count(neighbourhood_group)
```

- Question: In what way is `count()` less useful than `summarize()`?
- Answer: It is less flexible. Summarize can do many things, like add columns, find the average, etc. Count only counts things.

## Task 06. Room types in the Bronx?

- Question: What type of room is most common in the Bronx?
- Answer: Private roon

Which way would you prefer to solve this? `dplyr` or `ggplot`

```{r}
## YOUR CODE HERE!
abnb |>
  filter(neighbourhood_group == "Bronx") |>
  count(room_type)
```

## Task 07. Room price in Belmont?

- Question: What is the average price of a room in Belmont?
- Answer: 77.125 

```{r}
## YOUR CODE HERE!
abnb |>
  filter(neighbourhood == "Belmont") |>
  summarize(avg_price = mean(price))
abnb |>
  filter(neighbourhood == "Belmont", room_type == "Private room") |>
  summarize(avg_price = mean(price))
```

## Task 08. Predictor of price

- Question: Do you think the `number_of_reviews` is a good predictor of `price`?
- Answer: 

Task: Create a scatter plot showing the relationship between the two variables. You will want to use `geom_point` in your plot. Please discuss and upload your graph.

```{r}
## YOUR CODE HERE!
abnb |>
  ggplot(aes(x = number_of_reviews, y = price)) +
  geom_point()
```

## Task 09. Host price

- Question: How would you describe the distribution of host prices?
- Answer: 

Task: Draw a density plot of host price and upload it to Canvas.

- How would you describe it? Is it normally distributed? Does it have a long tail? Is it skewed?
- You will need to use `geom_density()` or `geom_histogram()` to answer this question.
- For more info on [Skewness](https://en.wikipedia.org/wiki/Skewness), I recommend Wikipedia.

```{r}
## YOUR CODE HERE!
abnb |>
  ggplot(aes(x = price)) |>
  geom_density()
```

To download your plot, right click on the image and download it to your local computer so you can upload it to Canvas.

## Task 10. Room type outliers

- Question: Which type of room has the most outlier prices? Options:
    - "Entire home/apt" or
    - "Shared room"
- Answer: 

The price of Airbnb hosts is obviously all over the place, which is a fun way to say there are many outliers in price. Which `room_type` has the most outliers?

To answer this question, you can use a box plot (`geom_boxplot()`). If you'd like to read more about box plots, I recommend the [Wikipedia box plot](https://en.wikipedia.org/wiki/Box_plot) article.

```{r}
## YOUR CODE HERE!
abnb |>
  ggplot(aes(x = room_type, y = price)) +
  geom_boxplot()
```

STRETCH: Can you think of another way to answer this question that would not rely on a box plot?

```{r}
## YOUR CODE HERE!
abnb |>
  ggplot(aes(x = price)) |>
  geom_point()
```

## Task 11. Shared room price

- Question: What is the average price of of a "Shared room"?
- Answer: 

```{r}
## YOUR CODE HERE!
abnb |>
  filter(room_type == "Shared room") |>
  summarize(avg_price == mean(price))

abnb |>
  group_by(room_type) |>
  summarize(avg_price = mean(price))
```

## Task 12. Shared room count

- Question: How many "Shared room" entries are in the data?
- Answer: 1,160 

```{r}
## YOUR CODE HERE!
abnb |>
  filter(room_type == "Shared room")

abnb |>
  group_by(room_type) |>
  summarize(n = n())
```

## Task 13. Useful?

- Question: Is this a large enough sample to be useful?
- Answer: 


Nothing to code here. This is a thinking out loud question.

All else being equal, a data scientist would rather have more than less data available. But, collecting data isn't free and takes time. Do you think we have enough data about the price of "Shared Rooms" to properly estimate it's cost in NYC?

## Task 14. Predicting Price

- Question: Which features (columns) do you think would be most useful for predicting price?
- Answer: Their answer depends on how well the support it. 

Soon we will start thinking about building a linear model to predict Airbnb host price in NYC. Use your transformation/plotting skills to look at the data. Describe what you find in Canvas. You can choose to use data transformation(s), plot(s), or a combination of both. Feel free to create additional code chunks if that is helpful to you.


# Next steps

Repetition is the key to learning this stuff. We are going to spend couple of weeks doing this over and over and over again.
