--
title: "Munge"
author: "Intro to Data Science"
date: "Week 03"
format:
  html:
    embed-resources: true
---


# Goals

- Lecture 01: Goals/Setup
- Lecture 02: Ggplot Review
- Lecture 03: Intro To Data Munging <- This is where we will start in class.
- Lecture 04: `filter()`
- Lecture 05: `count()`
- Lecture 06: `summarize()`
- Lecture 07: `group_by()`
- Lecture 08: THE PIPE
- Lecture 09: `arrange()`
- Lecture 10: Conclusions
- I think 10 videos is enough for one week!

Most of these will be SHORT lectures. I'm cutting them up to make them more accessible.


# Setup

- All of the lectures will be based on `lecture.qmd` and are based on a single session.
- How I organize RStudio Desktop.
- Tonight's data: Fuel Economy 2020
    - The US Federal Government publishes annual estimates for the mileage of cars sold in the US each year.
    - Raw Source: [https://www.fueleconomy.gov/](https://www.fueleconomy.gov/) 
    - You do not need to download this data. I did that for you and I added it here.
    - Review the [script I wrote](./data/get_mpg.R).
    - If you want to look yourself, at the bottom of the page, "Download EPA's MPG Ratings"
- The data extract we will use is in the `data` folder.

```{r}
#| label: setup
#| message: false
#| warnding: false

## Experiment with what happens if you set message to TRUE.
library(knitr)
library(rio)
library(tidyverse)
mpg <- import("data/mpg2020.csv")
```

Different ways to review the data:

1. See the data in the notebook as a table.
2. See the data in the notebook via `glimpse()`.
3. View the data in RStudio.

The first two options keep the data view in the notebook. The last option separates the data view from the notebook which reduces the value of the notebook as a stand-alone document. But, it is very flexible/useful.

## Inline (Notebook)
```{r}
## This gets awkward for wide data sets.
mpg
```

## Separate Tab

```{r}
## Uppercase "V" here is important!
View(mpg)
```


# Lecture 02: Ggplot Review

I hope you all find this graphic useful.

```{r}
#| echo: false
#| fig-align: 'center'

file <- "includes/ggplot flow.drawio.png"
knitr::include_graphics(file)
```

## Grain of `mpg`

- The book uses the term observation. I tend to use the data modeling term, grain.
- They mean the same thing.
- The grain of the `mpg` data is car _model_ from 2020.
    - Each row is a different model.
    - Brands span many rows.
    - We have data about the car model, but not about individual cars.
- So what?
    - This data tells us something about the mileage of car models for sale.
    - It tells us nothing about the mileage of cars actually sold.
    - It tells us nothing, really, of mileage of cars on the road.
    - [2020 Truck, SUV, Car Sales: Winners And Losers](https://www.forbes.com/wheels/news/2020-truck-suv-car-sales-winners-and-losers)
    
For example:

```{r}
#| echo: false

mpg |>
  filter(MfrName == "General Motors")
```

- One line, one observation, one car model sold, but not the number of times sold.
- Cannot calculate the (Corporate Average Fuel Economy (CAFE))[https://en.wikipedia.org/wiki/Corporate_average_fuel_economy#Calculation]

To have a functional `ggplot` you need TWO commands:

- ggplot
- geom_*

Below, is a SIMPLE example. Using this week's data, this plot shows that there is a pretty strong correlation between "City" and "Highway" gas mileage.

```{r}
ggplot(data = mpg, aes(x = City, y = Hwy)) +
  geom_point()
```

- Above, we can see that if we have two commands, we get a plot.
- But not all plots are so simple.
- The graph below shows us HOW MANY models there are in our data set, per manufacturer.
    - This has NOTHING to do with how many cars are actually SOLD.
    - In 2020 Ford and Toyota each sold NEARLY two million cars each.
    - Chevrolet and Honda each sold more than one million cars each.
    - GMC sold a little over 500K.
    - Again, this is where UNDERSTANDING THE GRAIN is important.

```{r}
ggplot(data = mpg, aes(x = MfrName)) +
  geom_bar(fill = "darkred") +
  labs(x = "Manufacturer", y = "Count", title = "Number of Models Sold") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

- What if we want to show percents instead of raw numbers? 
- Well, one way to do so . . . . there is an easier way we will discuss later.

```{r}
ggplot(data = mpg, aes(x = MfrName)) +
  geom_bar(aes(y = 100 * (..count..) / sum(..count..)), fill = "darkred") +
  ylim(0, 15) +
  labs(x = "Manufacturer", y = "Percent") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

- This plot aggregates (rolls up) a series of individual rows and GROUPS the results by manufacturer.
- This is data aggregation or summarization.
- This code should look familiar  ... although I added a couple of things:
    - `geom_bar(aes(y = 100 * (..count..) / sum(..count..)), fill = "darkred") +` This calculates percents rather than raw counts.
    - `ylim(0, 100) +` And this forces R to give us the full percent range.
    - `theme(axis.text.x = element_text(angle = 45, hjust = 1))` This protects our eyes from too many horizontal labels in a tight place.
- This plot shows PERCENTS, not COUNTS. For percents, it is best practice to show the full range of possible percents (0 - 100). Be very careful looking at a graph showing percents that doesn't show you where 0 and 100 are relative to the data shown. If they don't show you that, assume you are being manipulated (or the analyst is a amateur).
- If graphing data is your jam (as it should be!), you will enjoy looking through the [R Graph Gallery](https://www.r-graph-gallery.com/index.html). I would be remiss if I didn't casually mention that if you click through on the examples, the code to draw them is right there for you to read/learn from.
- Or we could just:

```{r}
ggplot(data = mpg, aes(x = MfrName)) +
  geom_bar(aes(y = 100 * (..count..) / sum(..count..)), fill = "darkred") +
  ylim(0, 20) +
  labs(x = "Manufacturer", y = "Percent") +
  coord_flip()
```

In this case, I actually like the sideways text, but that is for purely aesthetic reasons.


# Lecture 03: Intro to Data Munging

- Data Munging == Data Transformation
- The next few videos introduce a lot of new ideas.
- We transform data using functions (verbs) from the dplyr package (part of the Tidyverse). 
- Each function performs a specific data transformation.
- Why would we want to transform our data? To use it to answer questions.
  - The data we have is rarely exactly what we need in order to answer the question.
  - Example: I used this to show you only car models made by General Motors.
  - We 

```{r}
#| echo: false

mpg |>
  filter(MfrName == "General Motors")
```


Verbs:

- `filter()`: Only keep the rows of data that you need for the analysis. For example, we could use this function to drop all rows not made by Chevrolet.
- `count()`: Easy way to count the number of rows.
- `summarize()`: Provide summary statistics such as the average value of a column, or the number of lines in the data set.
- `group_by()`: Group, or stratify our results by a column. 
- `arrange()`: Sort, or arrange, the results by the values in column.

```{r}
url <- "https://upload.wikimedia.org/wikipedia/en/b/bd/ThePlaneteers.JPG"
knitr::include_graphics(url)
```

With their powers combined, we can perform many interesting transformations of our data, to answer questions the original data cannot answer. We will learn about other data transformations in the coming weeks. This is enough to get us started doing something interesting.

**BIG IDEA:** The data you have is ALMOST NEVER the data you want.

# Vaccine Hesitancy

- Measles was declared eliminated in the United States in 2000.
Eliminated means no endemic (internal) transmission of measles.
  - Cases still exist because measles exists in the world and someone could travel here and become symptomatic.
- There were more known cases of measles in 2025 than had been seen in decades.
- At this point, it is academic to argue about the status of measles in the US in 2026. Measles is endemic to the United States.
- The following table, from the CDC, troubles me.

|             | CY 2026 YTD | CY 2025 |
|:------------|------------:|--------:|
| Total Cases |         416 |    2255 |

| Age           | CY 2026 YTD | CY 2025   |
|:--------------|------------:|----------:|
| Under 5 years	|   102 (25%) | 577 (26%) |
| 5-19 years    |   252 (61%) | 993 (44%) |
| 20+ years	    |    42 (10%) | 670 (30%) |
| Age unknown   |    20  (5%) |  15  (1%) |

- Ninety three percent of the cases were in unvaccinated people.
- Seventy percent of the measles cases were in people 19 years old or younger.
- Yet, only 21.5% of the United States is younger than 18.

Why? How do we change this narrative?

# Lecture 04: Filter

Uses Boolean logic to filter our data.

- `==`
- `<`, `<=`, `>`, `>=`
- `is.na()`
- `!`

### How many car models were made by Chevrolet in 2020?

```{r}
## Data contains ONLY 2020 cars, so we don't have to filter by model year.
## Chevrolet is a "Division" of General Motors, so we have to filter that.
## We have to identify what data set we want to filter AND
## we have to tell R what Boolean test we want to use.
filter(mpg, Division == "Chevrolet")
```

In case you didn't realize this, General Motors owns/manufacturers Chevrolet and several car brands such as Cadillac & GMC.

- `Division` is a character column, so ONLY "==" makes any sense here.
- This wouldn't mean anything: `Division > "Chevrolet"`
    - Other than to point out how bad Chevies are.

### How many four cylinder model cars were made in 2020?

```{r}
## Notice how mpg is the first parameter.
filter(mpg, NCyl == 4)
```

### How many car models were made with fewer than 6 cylinders?

```{r}
filter(mpg, NCyl < 6)
```

In case you didn't realize this, there many different kinds of Honda Civic. We can find them all without having to know the `Carline` name for each!

```{r}
## This creates a new data set!
civics <- filter(mpg, str_detect(Carline, "CIVIC"))
not_civics <- filter(mpg, !str_detect(Carline, "CIVIC"))
```

Compare the city gas mileage of Civics to the rest of the cars.

```{r}
## It is OK to put two (or more) plots, etc. in a single code chunk.

## The distribution of city mileage for civics.
ggplot(civics, aes(x = City)) +
  geom_density(color = "darkgreen") +
  labs(title = "Civics")

## The distribution of city mileage for all.
ggplot(mpg, aes(x = City)) +
  geom_density(color = "darkred") +
  labs(title = "All")

## The distribution of city mileage for all.
ggplot(not_civics, aes(x = City)) +
  geom_density(color = "darkblue") +
  labs(title = "Not Civics")
```

Some takeaways:

1. How do you get a Honda Civic that can only get 22/23 mpg in the city?
2. PLEASE observer how these two graphs peak in very different places!
3. The first parameter in the filter command AND the ggplot command is data.
4. While this data does not allow us to control for the number of units shipped, it is still sobering to see that the average city gas mileage of cars sold in the US in 2020 was just less than 20 mpg. This is only marginally higher than the mileage on my 2013 Nissan Xterra.
5. The first function parameter in EVERY function we have used today has been a data set.


# Lecture 05: Count

- Count is a convenience function. It makes it easy, to count things.
- Do you remember this plot?

```{r}
ggplot(data = mpg, aes(x = MfrName)) +
  geom_bar(fill = "darkred") +
  labs(x = "Manufacturer", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

- It shows us the number of car models for each manufacturer. 
- And it is clear that GM has the most models.
- This plot summarizes the 1,164 rows of data into a graph with 26 columns.
- Can tell me exactly how many models Nissan has?
  - No? I didn't think so.

```{r}
count(mpg, MfrName)
```

- This summary table summarizes the 1,164 rows of data into a summary table with 26 rows.


# Lecture 06: Summarize

- Creates summary tables, like the function count, of a data set.
- If used by itself, it creates one row per data frame.

Uses helper functions to operate.

- `min()`, `max()`
- `mean()`, `sd()`
- `n()`

A summary statistic, hence summarize, is a single number that tells us something about the data in that column.

And we can see how this plays out when we run some summary statistics on the `civics` and `mpg` data sets.

```{r}
summarize(
  civics,
  N = n(),
  CityAvg = mean(City),
  CitySD = sd(City),
  CityMin = min(City),
  CityMax = max(City)
)

summarize(
  mpg,
  N = n(),
  CityAvg = mean(City),
  CitySD = sd(City),
  CityMin = min(City),
  CityMax = max(City)
)
```

Does this make sense, in the context of our data set and density plot above?


# Lecture 07: Group By

- Stratifies the results of a summarize.
- Function: `group_by()`
- That underscore is important. The function is not `group by()`!!!
- This is like our bar chart . . . . and our `count()` function.

```{r}
ggplot(data = mpg, aes(x = MfrName)) +
  geom_bar(fill = "darkred") +
  labs(x = "Manufacturer", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

count(mpg, MfrName)
```

```{r}
grouped_mpg <- group_by(mpg, MfrName)
summarize(grouped_mpg,
  N = n(),
  CityAvg = mean(City),
  CitySD = sd(City),
  CityMin = min(City),
  CityMax = max(City)
)
```



```{r}
grouped_civics <- group_by(civics, Carline)
summarize(grouped_civics,
  N = n(),
  CityAvg = mean(City),
  CitySD = sd(City),
  CityMin = min(City),
  CityMax = max(City)
)
```

- There's a five-door Civic?
- The function `group_by` creates a grouped data set which is visually indistinguishable from an un-grouped data set.
- This command is not useful on it's own.
- However, a grouped data set summarizes differently than it otherwise would.
- Again, notice how the data set is ALWAYS the first parameter.
    - There must be some reason I keep harping on this. . . .
  
```{r}
## The only differentiating feature is the part that says
## Groups: Carline [3]
grouped_civics
```

In truth, `group_by()` is a convenience function. We could, if we wanted write:

```{r}
## This just concatenates the results together into a single tibble.
bind_rows(
  ## This is where the real action is at.
  ## I'm doing the SAME calculation three times!
  ## This is hard to read and hard to maintain.
  summarize(
    filter(civics, Carline == "CIVIC 2DR"),
    Carline = "CIVIC 2DR",
    N = n(),
    CityAvg = mean(City),
    CitySD = sd(City),
    CityMin = min(City),
    CityMax = max(City)
  ),
  summarize(
    filter(civics, Carline == "CIVIC 4Dr"),
    Carline = "CIVIC 4Dr",
    N = n(),
    CityAvg = mean(City),
    CitySD = sd(City),
    CityMin = min(City),
    CityMax = max(City)
  ),
  summarize(
    filter(civics, Carline == "CIVIC 5DR"),
    Carline = "CIVIC 5DR",
    N = n(),
    CityAvg = mean(City),
    CitySD = sd(City),
    CityMin = min(City),
    CityMax = max(City)
  )
)

summarize(
  group_by(civics, Carline),
  N = n(),
  CityAvg = mean(City),
  CitySD = sd(City),
  CityMin = min(City),
  CityMax = max(City)
)
```

And now, I hope, you can better understand why `group_by()`, like `facet_wrap()`, exists. It doesn't do anything _new_, it just does it more easily.


# THE PIPE

Look back at that code above. It was nested. I had a `filter()` inside a `summarize()` inside a `bind_rows()`. It was hard to read.

Pipes (`|>`) offer a better way!

```{r}
## Replaces an explicit assignment of data, with an implicit assignment of data.
## The next two code chunks do the same thing.

## When using a pipe, the result set from the previous function is called ".".
## This is a convention. It could have been Bob, but it is called ".".
civics |>
  group_by(Carline) |>
  summarize(
    N = n(),
    CityAvg = mean(City),
    CitySD = sd(City),
    CityMin = min(City),
    CityMax = max(City)
  )
```

This is EXACTLY the same as what we did above, but we did it in a single step. The pipe symbol `|>` sends the output from one function to the input of the next function and works when the first parameter of the function is the data. That's why I kept pointing this out. We can do all sorts of piping!

If you see me type: `%>%`:

  - `|>` is the new R pipe.
  - If/ when you see the pipe from the magrittr package (`%>%`), it is basically the same thing. 
  - The only difference is that `|>` is part part of base R and `%>%` is part of a package. But the idea proved to be so popular they decided requiring a package for a pipe ought not be necessary.

### Using pipes, calculate the mean, sd, min, and max values of City mileage for Honda Civics.

- Worst title of the lecture goes to . . . .

```{r}
easy_summary_civics <-
  mpg |>
  filter(str_detect(Carline, "CIVIC")) |>
  group_by(Carline) |>
  summarize(
    N = n(),
    CityAvg = mean(City),
    CitySD = sd(City),
    CityMin = min(City),
    CityMax = max(City)
  )

easy_summary_civics
```

```{r}
ggplot(easy_summary_civics, aes(x = Carline, y = CityMax)) +
  geom_col()

ggplot(civics, aes(x = Carline)) +
  geom_bar()
```

Thanks to the pipe operator (`|>`), we can avoid creating a million data-step data sets. This is kinda-sorta available in SAS, but it is not taught as a best practice. In modern R programming, using the pipe is considered best practice and is how this stuff gets taught.


# Lecture 08: Arrange

After performing one or more data transformations, R will return our data to us in alphabetical order, based on the first column. For example, see above.

But, that isn't ALWAYS how we want it.

- Function: `arrange()`
- By default, returns are sorted in ascending order.
- `desc()` Returns can be sorted in descending order.

```{r}
easy_summary_civics |> arrange(CityAvg)
easy_summary_civics |> arrange(desc(CityAvg))
```

This function is a useful function on your labs!!!!!! For example, say I ask a question like this:

> Task X: Which car model in mpg has the HIGHEST highway gas mileage?

So, there are over 1,000 rows in mpg. I don't think you want to go searching through that by hand. In fact, I KNOW you don't. (If you do, let's talk after class.) Fortunately, arrange makes it easy.

```{r}
mpg |>
  arrange(desc(Hwy)) |>
  select(MfrName, Carline, Hwy)
```

## Combine ggplot and dplyr!

Task: What is the percent of cars by car class (`CarlineClassDesc`). Using this, what is the most common car class in this data set?

```{r}
mpg |>
  count(CarlineClassDesc) |>
  mutate(p = 100 * n / sum(n)) |>
  ## arrange(desc(p)) |>
  ggplot(aes(x = p, y = CarlineClassDesc)) +
  ## Don't use geom_col b/c of the grain of the data coming into ggplot.
  geom_col(fill = "darkred") +
  ## Please note what I did here. I capped the plot at 25%.
  xlim(0, 100) +
  labs(title = "The greatest plot in history", x = "Percent", y = "Car Class")
```

There's a lot going on here, but you've seen ALMOST all of it before.

1. We transform the data. Note how I am calculating both N counts and percentages.
2. We use `geom_col` rather than `geom_barm`.
3. With each step, we pass the results along the pipe. And at the end of the pipe, we get a plot.
4. THERE ARE TWO PIPES (sorta)
    - Data Transformations: `|>` (Pass it along this way >)
    - Data Viz: `+` (Add another layer)


# Conclusions

R lets you combine small functions flexibly to perform complex tasks.

```{r}
url <- "https://upload.wikimedia.org/wikipedia/en/b/bd/ThePlaneteers.JPG"
knitr::include_graphics(url)
```

## Use Data Responsibly 

Always look for documentation to help you understand the: 

- Content of the data (columns)
- Sampling/inclusion/exclusion criteria (as relevant/appropriate)
- How much data is there (this goes back to 1978)
- Caveats, changes in the data over time.
- Any aggregations we can use to validate our analyses.

For example, this data does NOT include EVERY vehicle sold in the US. From the PDF documentation on the EPA website, we learn that some vehicles are exempt.

>Light-duty fuel economy regulations do not apply to:
>
>- Sport utility vehicles (SUVs) and passenger vans with a gross vehicle weight rating (GVWR) of more than 10,000 pounds—GVWR is the vehicle weight plus carrying capacity
>- Other vehicles with a GVWR of 8,500 pounds or more or a curb weight over 6,000 pounds. Therefore, manufacturers do not have to estimate their fuel economy, and fuel economy labels are not posted on their windows. Also, fuel economy information on some vehicles was not available in time to be included in this guide. More up-to-date information can be found at fueleconomy.gov.

I lack the domain expertise to tell you if this is a good or bad decision. However, the fact remains that this data is incomplete.

https://www.caranddriver.com/features/a15388892/the-truth-about-epa-city-highway-mpg-estimates/

## Data Is Disappearing

> CDC’s website is being modified to comply with President Trump’s Executive Orders

- As of January (barely) 2026, Federal Agencies such as the CDC have been removed or limited the utility of many long-running data sets.
  - The administration's data deletion spree began in January 2025.
  - Executive Order 14168: Ending Radical and Wasteful Government DEI Programs and Preferencing
  - Executive Order 14151: Defending Women from Gender Ideology Extremism and Restoring Biological Truth to the Federal Government
- The deleted content included:
  - Public health information relating to LGBTQ health, 
  - Gender and reproductive health,
  - Vaccine guidance,
  - Mpox treatment,
  - pregnancy risk,
  - opioid use disorder,
  - HIV/AIDS research, and the NIH HIV Risk reduction tool,
  - data from clinical trials, and more
  - All data has limitations. This is censorship. This is an attempt to reframe the narrative. The loudest voice wins when there is no truth.
    - State-level crime data and data on LGBTQ hate crimes.
    - The National Law Enforcement Accountability Database, which tracked federal police officer misconduct.
    - Not One More Report on missing and and murdered Native Americans.
    - Youth Risk Behavior Surveillance System was later restored, but missing code books and information needed to use the data.
- September 2, 2025 A judge issued an order in Washington State Medical Association et al. v. Kennedy et al requiring DHS to restore the delted data because the Federal government had failed to follow the legal procedure for removing such resources.
  - Additional context: the court order requires the government to restore a few hundred data sets.
  - The government has remove over 8,000 websites and 3,000 data sets.
- Links:
  - [HHS Agrees to Settlement Requiring the Restoration of Deleted Health Data and Websites](https://www.hipaajournal.com/hhs-settlement-lawsuit-restore-critical-health-information-federal-websites/): September 10, 2025
  - [2025 United States government online resource removals](https://en.wikipedia.org/wiki/2025_United_States_government_online_resource_removals)
